Optimising Heterogeneous Ambulance Fleet Allocations in Jakarta

### 2025-01-21 response_2/response.tex

\documentclass{article}

\usepackage{fullpage}
\usepackage{parskip}
\usepackage{setspace}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{hyperref}

\title{Responses to Reviewers}
\author{}
\date{}

\begin{document}

\maketitle

We would like to express our gratitude to the editor, associate editor, and reviewers for facilitating the review process, and for their useful suggestions and remarks on the manuscript.

This document will address the specific concerns raised.

\section{Responses to the Editor}

There were no comments from the Editor.

\section{Responses to the Associate Editor}
\begin{quote}
``Thank you for your revised manuscript. While one reviewer is satisfied with
    the changes, the second reviewer has raised additional concerns that require
    further clarification. It is clear that you have addressed many of the
    previous remarks; however, some points still need more detailed responses.
    Please provide a clear and thorough explanation for the points raised by the
    second reviewer in your next revision. We look forward to your updated
    manuscript.''
\end{quote}

We are delighted to see that we have addressed all the points raised by reviewer
2. Through this response we will endeavour to demonstrate clearly and precisely
how we have addressed the remaining points raised by reviewer 1.

\section{Responses to the Reviewers}

\subsection{Comments from Reviewer 1}

\begin{itemize}

\item Reviewer 1's comment from the first review:

\begin{quote}
``No measures of variance are given for the simulation results in Tables 1
and 4. Especially since repeated simulation results were conducted, measu-
res of variance should be included to investigate the statistical significance
of the results.''
\end{quote}

\item Our response to reviewer 1's comment from the first review:

Thank you for the comment, standard deviations of mean response times are now
given in Table 1 and 4. This is a particularly useful inclusion as the
large variance in the response times explains the discrepencies in the
simulated and expected survivals.

\item Review 1's new comment:

\begin{quote}
Not addressed: The standard deviation of the response time was inclu-
ded, but no information about the variance of the results over the eight
replications was given. To obtain knowledge about the reliability of the
simulation results, some information should be given about how much the
simulation results differed between the eight replications mentioned on
page 17.
\end{quote}

A short discussion about the variability over the repetitions (including
variance) has now been included. 
% TODO VK: We need to do this, I believe it is only Geraint who has the data.

\item Reviewer 1's comment from the first review:

\begin{quote}
``Unclear, why the subchapter 4.1 on survival functions is necessary, sin-
ce traditional coverage targets are used in the case study and the model
formulation does not depend on characteristics of the survival functions.
While survival functions are an important performance metric in the re-
viewers’ opinion, it is not clear why the reader needs this information to
follow the remainder of the pape.''
\end{quote}

\item Our response to reviewer 1's comment from the first review:

We disagree with the reviewer here. In the case study, traditional targets such
a mean response times are reported alongside survival measures, based on
the survival functions. However, a key part of the paper is the
optimisation process, and all scenarios compared are `optimised'
scenarios that result from this process. The objective function of the
optimisation process maximises survival, based on survival functions.
This is evnidenced in Equations 6 and 7, which are component parts of
the objective function given in Equation 5. Therefore the survival
functions are central to the work presented.

\item Review 1's new comment:

\begin{quote}
''Not addressed: It was expanded upon why survival functions can be sui-
table performance metrics. In the case study however, coverage targets
were used (8 min target for A1 patients, 15 min target for A2 patients
and 60 min target for B patients). It is therefore still unclear, why the
reader needs the information in chapter 4.1. Additionally, there is no dis-
cussion as to why coverage targets were chosen in the case study, after
it was argued, that survival functions can solve some drawbacks of using
coverage targets.``
\end{quote}

and

\begin{quote}
    ``
There is contradicting information about which survival functions were
used in the case study. On page 15 and 16, response time targets for the
different categories are listed. This implies to the reader, that step-wise
survival functions, in other words coverage targets, were used in the case
study. On page 19, line 44 it is stated, that Equation 3 was used as the
survival function for A1 patients. It needs to be clarified, which survival
functions were used in the MESLMHPHF in the case study both for the
optimization metaheuristic and the evaluation of the results.
    ''
\end{quote}

\item Our response:

We have added a sentence to the text to clarify that despite the policy target
of 8 minutes for the A1 patients, we have in fact used the survival function of
equation (3) for A1 patients. This was not explicit and we appreciate the
reviewer insisting this point.

\item Reviewer 1's comment:

\begin{quote}
``The simulated survival probabilities in Table 4 show, that the survival for
A1 patients was not improved in any of the four scenarios in the simu-
lation. As the simulation model is a more realistic model of the actual
1
EMS system in comparison to the MESLMHPHF, the simulated survi-
val probabilities carry more weight in interpreting the results compared
to the expected survival probabilities calculated with the MESLMHPHF.
On page 19, line 42 - 44 however it is stated, that ‘in particular A1 pati-
ents are benefitting from the improved allocations’. It is unclear, how the
authors draw this conclusion, when the simulation results show worse or
equal survival probabilities for A1 patients in all scenarios.''
\end{quote}

The reviewer raises a good point here. We have modified the text to hopefully
clarify that:

\begin{itemize}
    \item The sentence regard A1 patients meant to reflect more on the
        efficiency and accuracy of the optimisation heuristic which we find is
        one of the main contributions of this paper (specifically the novel
        approach used to deal with the circular nature of utilisation and
        allocation).
    \item The optimised allocation give more robust allocations under increasing
        demand although we admit that there is not enough numerical evidence to
        be sure of this effect.
\end{itemize}

Furthermore we have more explicitly referred to our explanation about the cause
of this discrepancy.

Fundamentally, again, the reviewer raises a valid and interesting avenue for investigation and
one that we feel
could not be precisely considered without the work in this paper. We hope to
tackle building a more accurate objective function, that better captures the
stochastic effects, that could be then optimised with our novel algorithm. This,
does not however fit in this paper. If the reviewer would like us to make that
more explicit and clear in the text we would be happy to follow any guidance
they feel is necessary.

\item Reviewer 1's comment:

    \begin{quote}
        ``      Equations 8 and 9: It is stated, that \(\beta_{pa_1a_2}\) indicates, that a vehicle from
        \(a_1\) can reach \(p\) quicker, than a vehicle from \(a_2\). The equations 8 and 9
        however use \(\leq\) and not \(<\) , which is not consistent with the description.
Has been fixed for \(\beta_{pa_1a_2}\) in Equation 8, but not for \(R_{pa_1a_2}\) in Equation
9. ''
    \end{quote}
    This has been fixed.
    % TODO Check with Geraint that this is correct.

\end{itemize}

\subsection{Comments from Reviewer 2}

There were no further comments from Reviewer 2.

\end{document}

### 2025-01-21 tex/main.tex
Fix column name.

The application of our research is focused on Indonesia, supported by an
award for Global Challenges Research Funds (GCRF). In Indonesia, an LMIC,
there has been very little prior work on developing an EMS strategy, driven
by \textcolor{red}{\st{inadequate}} a lack of financial investment and human capital
\citep{plummer2017ems,pusponegoro2003terrorism,yusvirazi2018state}.
Related research by the authors \citep{BriceSyaribahNoor2022Esui} has
highlighted particular challenges in Indonesia as a lack of a single
coordinated EMS, the ability or willingness to pay for an ambulance, a large
geographical area, and areas of severe traffic congestion especially in
Jakarta, the capital.

@@

\textcolor{red}{There are several managerial implications for EMS providers and
government agencies in Jakarta and Indonesia, respectively. The findings
emphasise the importance of adopting a dynamic and data-driven approach to
resource allocation. The inclusion of a heterogeneous fleet, such as Rapid
Response Vehicles (RRVs) alongside traditional ambulances, has been shown to
significantly improve response times and patient survival rates, particularly in
densely populated urban areas. Managers are advised to invest in these diverse
fleets and adopt flexible allocation strategies that account for temporal
variations in demand, such as the daily influx of commuters and the differing
population densities across neighbourhoods. The research underscores the
necessity of enhancing data collection practices to support informed
decision-making, with the development of user-friendly decision support tools to
assist in visualising data and predicting outcomes}

@@

\begin{itemize}
  \item $\lambda_{pk}$ be the rate at which patients of speciality $k$ make
        calls from pick-up location $p$;
  \item $s_k(t)$ be the survival function function associated with patients of
        speciality $k$;
  \item $\pi_a$ be the average utilisation of primary vehicles stationed at
        location $a$;
  \item $b_{pa}$ be the expected travel time from ambulance location $a$ to
        pick-up location $p$; and let $B_{pa}$ be a random variable representing
        the time it takes for an ambulance to drive this distance;
  \item $C_{py}$ be a random variable representing the travel time from pick-up
        location $p$ to hospital $y$;
  \item $D_{ya}$ be a random variable representing the travel time from hospital
        $y$ to ambulance location $a$;
  \item $F_{pa}$ be a random variable representing the travel time from pick-up
        location $p$ to ambulance location $a$;
  \item $G_k$ be a random variable representing the time the ambulance spends
        with patients of speciality $k$ at the pick-up location;
  \item $J_k$ be a random variable representing the time the ambulance spends
        with patients of speciality $k$ at the hospital;
  \item $\Theta$ be a random variable representing the time the ambulance spends
        re-fuelling, re-stocking, and resting between transit jobs;
  \item $q_{pky}$ be the probability that a patient of speciality $k$ from
        pick-up location $p$ is taken to hospital $y$.
        Note that \textcolor{red}{having} $\sum_{y \in \mathcal{Y}} q_{pky} < 1$ is possible, that is a
        patient may not go to any hospital, in which case the ambulance returns
        to their ambulance location.
\end{itemize}

@@

\textcolor{red}{Finding allocations that maximise survival can result in
allocations with different behaviours to those found by minimising response
times or maximising coverage. Coverage maximisation models are often easier to
solve, and easier to communicate, but may not take into consideration severity
or intensity of demand \citep{Erkut200842}. This results in a minority of
patients, particularly those more isolated, being over-serviced, that is, having
more relative resources allocated to their needs that the majority, at the
expense of the majority. Minimising average response times on the other hand
treats all patients equally, but may have the unintended outcome of disregarding
some more further out patients whose response times would be larger that others
regardless of the allocation. It would still disregard the severity or type of
the patient. Allocations based on maximising survival are similar to those that
minimise response time, overcoming the issue of over-servicing outlying
patients, though it is still possible to disregard them completely. It has an
advantage over minimising response times as it can account for patient severity.
Having survival as functions of response times has the effect of weighting gains
in response times by how valuable those gains are to a patient.}

\subsection{Maximal Expected Survival Location Model with Heterogeneous
            Patients and Heterogeneous Fleet}\label{sec:meslmhphf}
Here we propose a Maximal Expected Survival Location Model with Heterogeneous
Patients and Heterogeneous Fleet (MESLMHPHF). It is an extension of the
MESLMHP model given in \citet{Knight2012918}, which did not consider
heterogeneous fleets. This is a model of survival that can be used as an
objective function for optimisation algorithms. It is a weighted expected
survival function as the objective function which considers heterogeneous
patients (different specialities) and heterogeneous fleets (both primary
vehicles, EAs, and secondary vehicles, RRVs), and is constructed by
appropriately summing these survival curves, described in
Section~\ref{sec:survival}, across the population, multiplying by ambulance
availability where appropriate.
\textcolor{red}{The full MESLMHPHF optimisation model is given in
APPENDIX~\ref{apx:fullmodel}, and its components are described fully here.}
The MESLMHPHF is given by Equation~\ref{eqn:meslmhphf}.

@@

Here $R_{p a_1 a_2}$ and $\beta_{p a_1 a_2}$ \textcolor{red}{define an ordering
of the ambulance locations, they} are binary variables indicating the preference
of sending a vehicle from one station to another, that is the dispatch rule.
Here we use the closest or fastest vehicle, so $\beta_{p a_1 a_2}$ indicates if
a vehicle of the same type can reach $p$ quicker from $a_1$ than $a_2$, defined
in Equation~\ref{eqn:beta}, while $R_{p a_1 a_2}$ indicates if a primary vehicle
at $a_1$ can reach $p$ quicker than a secondary vehicle at $a_2$, defined in
Equation~\ref{eqn:R}. \textcolor{red}{In order for these to define an ordering
it is assumed that no two distinct ambulance locations are exactly equidistant
from a pickup location.} Note that here these parameters are defined using
travel times, but can be generalised to account for any type of preference,
including for example proximity, importance, or efficiency.

@@

Equation~\ref{eqn:survival_A} is the probability of a patient surviving (their
survival function), multiplied by the availability of primary vehicles at that
ambulance station, multiplied by the unavailability of vehicles at closer
stations.
Equation~\ref{eqn:survival_B} extends this to two vehicles types, the first
part of the sum repeating the logic for the faster secondary vehicles, and the
second part of the sum adapting that logic for primary vehicles, who will only
be contribute the patient's survival if all faster secondary vehicles are also
busy. These interpretations are given as annotations to the equations
in~\textcolor{red}{APPENDIX}\ref{apx:annotated}.

\textcolor{red}{Note that in Equations~\ref{eqn:survival_A}
and~\ref{eqn:survival_B}, the survival functions used, $s_k(t)$, are the same
whether a primary or a secondary vehicle arrives first. This may be unrealistic,
where primary vehicles may be better equipped than secondary vehicles to deal
with some emergencies. One option to overcome this would be to have distinct
survival functions for each vehicle type, however there may be some interactions
between the survival probability functions that would not be captured, for
example when a primary vehicle arrives after a secondary vehicle, there may
still be a better chance of survival that the secondary vehicle only. This would
be an interesting avenue of further research. For this work, survival functions
will be the same for all vehicles, representing the probability of surviving
given time to be seen by any emergency vehicle.
Note also that the survival functions are functions of $b_{pa}$, the expected
travel time between $p$ and $a$, and do not consider the variability in the
travel times. This will be discussed further alongside results in
Section~\ref{sec:compare_exp_and_sim}.}

@@

\noindent
then, using the simulation described in Section~\ref{sec:simulation}, we
consider the distribution of $\rho$ across all simulated jobs under the current
situation and allocation in Jakarta, shown in Figure~\ref{fig:rho_distribution}.
We see for primary vehicle jobs on average 24\% of a job time is location
dependent, while it is 36.5\% for secondary vehicles. Considering
location-dependent service rates would be an interesting further research
question, potentially improving the model's performance by incorporating travel
times into the MESLMHPHF. However, given that in the case of Jakarta the
preferences $\beta_{p a_1 a_2}$ and $R_{p a_1 a_2}$ are themselves defined by
travel times (Equations~\ref{eqn:beta} and~\ref{eqn:R}), this \textcolor{red}{might not have a strong impact on the results of \st{may not have much
affect for}} this study, \textcolor{red}{as the model already gives preference to
the locations with the shortest service times.}

@@

We propose finding the true vehicle utilisations by first solving
Equation~\ref{eqn:utilisations_primary} for the $\lambda_a$; determining the
primary vehicle utilisations with Equation~\ref{eqn:utilisation_ratio_primary};
then using these to solve Equation~\ref{eqn:utilisations_secondary} for the
$\tilde{\lambda}_a$; and determining the secondary vehicle utilisations with
Equation~\ref{eqn:utilisation_ratio_secondary}. These can be solved
numerically. In our implementation this is solved using the MINPACK hybrd and
hybrj algorithms by using Scipy's \texttt{fsolve} function
\citep{2020SciPy-NMeth}. \textcolor{red}{These algorithms are efficient methods for finding zeros of systems of non-linear equations. They are based on the Powell hybrid method \citep{powell70}.} It should be noted that in this section we assume that
all $\lambda_{pk}$ are static and not time-dependant, which is an assumption
that will be used throughout the optimisation methodology, however in the
simulation we can account for time-dependent demand.

@@

\begin{algorithm}
\algnewcommand\algorithmicto{\textbf{to }}
\caption{Evolutionary Algorithm used to find better allocations.}
\label{alg:heuristic}
\begin{algorithmic}[1]
\State Create a population of $N$ random allocations ($Z_a$, $\tilde{Z}_a$)\;
\For{$i \leftarrow 0$ \algorithmicto $H$}
  \State Rank the population \textcolor{red}{by decreasing expected survival} according to Equation~\ref{eqn:meslmhphf}\;
  \State Keep the top $\kappa$ solutions\;
  \State $m \leftarrow \lceil m_0 - \min\left(m_0, i \times c\right)\rceil$\;
  \For{$j \leftarrow 0$ \algorithmicto $N - \kappa$}
    \State Randomly choose and copy a solution from the population\;
    \State Mutate that copy of the solution $m$ times\;
    \State Place mutated copy in a separate population\;
  \EndFor
  \State Combine populations;\
\EndFor
\State Rank the population according to Equation~\ref{eqn:meslmhphf};\
\State Output the top ranking solution.
\end{algorithmic}
\end{algorithm}

@@

This is the method used to calculate the expected values of $B_{pa}$, $C_{py}$,
$D_{ya}$ and $F_{pa}$.
Furthermore it is assumed that travel times for each segment of the route
(from vehicle location to patient, from patient to hospital, and from hospital
back the the vehicle's location) follow a Triangular distribution around the
calculated expected travel time, between 75\% and 125\% of the value.
\textcolor{red}{A Triangular distribution is a useful distribution to use when
not much is known about the travel times except some basic summary statistics
\citep{robinson2014simulation}, in this case a likely range of values.}

@@

This is possible as there are no synchronicity issues between the two
components. Primary vehicles operate independently of secondary vehicles, that
is\textcolor{red}{,} the way primary vehicles respond to a patient is unaffected by the presence
or lack of a secondary vehicle.
Therefore, primary vehicle logic is not compromised by simulating primary
vehicles in isolation.
Secondary vehicles are impacted by the behaviour of primary vehicles, they
must remain with the patient until the primary vehicle arrives, and so must be
simulated after the simulation of primary vehicles, taking as inputs the exact
list of events that occurred. That is\textcolor{red}{,} the logic of the secondary vehicles is
determined by observing the actions of primary vehicles and reacting to them.
Simulation results are combined for each individual patient, to determine
their response time caused by either primary or secondary vehicles.

@@

Exact synchronicity considerations are shown in Figure~\ref{fig:sequential_logic}:
\begin{itemize}
  \item whenever the secondary vehicle reaches the patient before the primary
  has left, vehicle, they remain with the patient until the primary vehicle
  leaves;
  \item if secondary vehicle arrives after the primary vehicle has left, they
  will immediately return to their stations as they are not needed at the scene;
  \item if the expected time for the secondary vehicle to reach the patient
  exceeds the expected time for the primary vehicle to reach the patient then
  the secondary vehicle is not deployed, and so that transit job would be
  abandoned in the secondary simulation (although still seen by a primary
  vehicle in the initial simulation). We assume that secondary vehicles must
  still reach the pickup locations, due to potential difficulties in
  communication en-route. \textcolor{red}{Consider a situation when a
  secondary vehicle arrives and the patient does not require transportation to
  a hospital, with better en-route communication this would mean the primary
  vehicle turns around, however without en-route communication which is
  assumed here, the primary vehicle must still reach the patient pick-up
  location.}
\end{itemize}

@@

\textcolor{red}{
Verification of a model is ensuring that the conceptual model is implemented
correctly and without error, while validation of a model ensures that the
conceptual model accurately represents the real system to be modelled
\citep{robinson2014simulation}. Verifying the simulation code, alongside
verifying the optimisation and MESLMHPHF objective function code, is done
through automated code testing \citep{percival2014test, palmer2023implementing}.
This includes unit testing to ensure that each sub-component works are requires,
and is bug and error free; end to end testing, ensuring that the components
interact together as intended; and extreme value testing ensuring that changes
in input parameters result in sensible and expected changes in output results
and behaviours. The testing suites are included in the open code repositories
provided.
}

@@

\begin{itemize}
  \item \textbf{A1 - High priority emergency patients}: critical patients who
  require immediate life-saving assistance with a target ambulance response
  time of 8 minutes. In our data set there were 168 identified calls that met
  this criterion.
  \item \textbf{A2 - Other emergency patients}: urgent patients that require
  assistance with a target response time of 15 minutes. In our data set there
  were 23,784 calls of this type.
  \item \textbf{B - Non-emergency patients}: patients that still benefit from
  an ambulance response but non-critical with a target response time of 60
  minutes. There were 31,725 calls of this type in the data set.
\end{itemize}

@@

All models are parameterised using demand data that covers all 261
neighbourhoods from 1 January to 31 December 2019, before the COVID-19
pandemic. Data from 2020-2021 was naturally heavily skewed by the pandemic,
with significantly lower demand, and so was not considered as representative
of a typical period for forecasting future needs, hence the decision to use
the available 2019 demand. \textcolor{red}{APPENDIX}~\ref{apx:parameters}
provides further details on the parameterisation of the model, including how
travel time estimates were obtained.
\textcolor{red}{Verification of the models were discussed in
Section~\ref{sec:combined_model}, we validated based on expert opinion from
our collaborators, Ambulans 118 and the 119 Emergency Ambulance Service in
Jakarta, where several workshops have been carried out in Indonesia where the
model was shared and discussed. This expertise was vital in giving a basic
validation of the model, with collaborators approving model results
parametrised from their provided data and expert opinion.}

@@

Running the simulation and MESLMHPHF objective function of expected survival for
both the current and the two proposed grid allocations gives the results shown
in Table~\ref{tbl:current_grid_results}. The simulation was run over half a
year, with a month warm up time, over eight replications. We observe that the
two proposed grid allocations perform much worse than the current allocation as
measured by mean response time, survival, and percentage of calls abandoned.
This is expected as vehicle numbers are lower under the new proposals, as well
as the allocations focussing on coverage rather than response times or survival.

>>>>

The application of our research is focused on Indonesia, supported by an
award for Global Challenges Research Funds (GCRF). In Indonesia, an LMIC,
there has been very little prior work on developing an EMS strategy, driven
by a lack of financial investment and human capital
\citep{plummer2017ems,pusponegoro2003terrorism,yusvirazi2018state}.
Related research by the authors \citep{BriceSyaribahNoor2022Esui} has
highlighted particular challenges in Indonesia as a lack of a single
coordinated EMS, the ability or willingness to pay for an ambulance, a large
geographical area, and areas of severe traffic congestion especially in
Jakarta, the capital.

@@

There are several managerial implications for EMS providers and
government agencies in Jakarta and Indonesia, respectively. The findings
emphasise the importance of adopting a dynamic and data-driven approach to
resource allocation. The inclusion of a heterogeneous fleet, such as Rapid
Response Vehicles (RRVs) alongside traditional ambulances, has been shown to
significantly improve response times and patient survival rates, particularly in
densely populated urban areas. Managers are advised to invest in these diverse
fleets and adopt flexible allocation strategies that account for temporal
variations in demand, such as the daily influx of commuters and the differing
population densities across neighbourhoods. The research underscores the
necessity of enhancing data collection practices to support informed
decision-making, with the development of user-friendly decision support tools to
assist in visualising data and predicting outcomes.

@@

\begin{itemize}
  \item $\lambda_{pk}$ be the rate at which patients of speciality $k$ make
        calls from pick-up location $p$;
  \item $s_k(t)$ be the survival function function associated with patients of
        speciality $k$;
  \item $\pi_a$ be the average utilisation of primary vehicles stationed at
        location $a$;
  \item $b_{pa}$ be the expected travel time from ambulance location $a$ to
        pick-up location $p$; and let $B_{pa}$ be a random variable representing
        the time it takes for an ambulance to drive this distance;
  \item $C_{py}$ be a random variable representing the travel time from pick-up
        location $p$ to hospital $y$;
  \item $D_{ya}$ be a random variable representing the travel time from hospital
        $y$ to ambulance location $a$;
  \item $F_{pa}$ be a random variable representing the travel time from pick-up
        location $p$ to ambulance location $a$;
  \item $G_k$ be a random variable representing the time the ambulance spends
        with patients of speciality $k$ at the pick-up location;
  \item $J_k$ be a random variable representing the time the ambulance spends
        with patients of speciality $k$ at the hospital;
  \item $\Theta$ be a random variable representing the time the ambulance spends
        re-fuelling, re-stocking, and resting between transit jobs;
  \item $q_{pky}$ be the probability that a patient of speciality $k$ from
        pick-up location $p$ is taken to hospital $y$.
        Note that having $\sum_{y \in \mathcal{Y}} q_{pky} < 1$ is possible, that is a
        patient may not go to any hospital, in which case the ambulance returns
        to their ambulance location.
\end{itemize}

@@

Finding allocations that maximise survival can result in
allocations with different behaviours to those found by minimising response
times or maximising coverage. Coverage maximisation models are often easier to
solve, and easier to communicate, but may not take into consideration severity
or intensity of demand \citep{Erkut200842}. This results in a minority of
patients, particularly those more isolated, being over-serviced, that is, having
more relative resources allocated to their needs that the majority, at the
expense of the majority. Minimising average response times on the other hand
treats all patients equally, but may have the unintended outcome of disregarding
some more further out patients whose response times would be larger that others
regardless of the allocation. It would still disregard the severity or type of
the patient. Allocations based on maximising survival are similar to those that
minimise response time, overcoming the issue of over-servicing outlying
patients, though it is still possible to disregard them completely. It has an
advantage over minimising response times as it can account for patient severity.
Having survival as functions of response times has the effect of weighting gains
in response times by how valuable those gains are to a patient.

\subsection{Maximal Expected Survival Location Model with Heterogeneous
            Patients and Heterogeneous Fleet}\label{sec:meslmhphf}
Here we propose a Maximal Expected Survival Location Model with Heterogeneous
Patients and Heterogeneous Fleet (MESLMHPHF). It is an extension of the
MESLMHP model given in \citet{Knight2012918}, which did not consider
heterogeneous fleets. This is a model of survival that can be used as an
objective function for optimisation algorithms. It is a weighted expected
survival function as the objective function which considers heterogeneous
patients (different specialities) and heterogeneous fleets (both primary
vehicles, EAs, and secondary vehicles, RRVs), and is constructed by
appropriately summing these survival curves, described in
Section~\ref{sec:survival}, across the population, multiplying by ambulance
availability where appropriate.
The full MESLMHPHF optimisation model is given in
APPENDIX~\ref{apx:fullmodel}, and its components are described fully here.
The MESLMHPHF is given by Equation~\ref{eqn:meslmhphf}.

@@

Here $R_{p a_1 a_2}$ and $\beta_{p a_1 a_2}$ define an ordering
of the ambulance locations, they are binary variables indicating the preference
of sending a vehicle from one station to another, that is the dispatch rule.
Here we use the closest or fastest vehicle, so $\beta_{p a_1 a_2}$ indicates if
a vehicle of the same type can reach $p$ quicker from $a_1$ than $a_2$, defined
in Equation~\ref{eqn:beta}, while $R_{p a_1 a_2}$ indicates if a primary vehicle
at $a_1$ can reach $p$ quicker than a secondary vehicle at $a_2$, defined in
Equation~\ref{eqn:R}. In order for these to define an ordering
it is assumed that no two distinct ambulance locations are exactly equidistant
from a pickup location. Note that here these parameters are defined using
travel times, but can be generalised to account for any type of preference,
including for example proximity, importance, or efficiency.

@@

Equation~\ref{eqn:survival_A} is the probability of a patient surviving (their
survival function), multiplied by the availability of primary vehicles at that
ambulance station, multiplied by the unavailability of vehicles at closer
stations.
Equation~\ref{eqn:survival_B} extends this to two vehicles types, the first
part of the sum repeating the logic for the faster secondary vehicles, and the
second part of the sum adapting that logic for primary vehicles, who will only
be contribute the patient's survival if all faster secondary vehicles are also
busy. These interpretations are given as annotations to the equations
in APPENDIX~\ref{apx:annotated}.

Note that in Equations~\ref{eqn:survival_A}
and~\ref{eqn:survival_B}, the survival functions used, $s_k(t)$, are the same
whether a primary or a secondary vehicle arrives first. This may be unrealistic,
where primary vehicles may be better equipped than secondary vehicles to deal
with some emergencies. One option to overcome this would be to have distinct
survival functions for each vehicle type, however there may be some interactions
between the survival probability functions that would not be captured, for
example when a primary vehicle arrives after a secondary vehicle, there may
still be a better chance of survival that the secondary vehicle only. This would
be an interesting avenue of further research. For this work, survival functions
will be the same for all vehicles, representing the probability of surviving
given time to be seen by any emergency vehicle.
Note also that the survival functions are functions of $b_{pa}$, the expected
travel time between $p$ and $a$, and do not consider the variability in the
travel times. This will be discussed further alongside results in
Section~\ref{sec:compare_exp_and_sim}.

@@

\noindent
then, using the simulation described in Section~\ref{sec:simulation}, we
consider the distribution of $\rho$ across all simulated jobs under the current
situation and allocation in Jakarta, shown in Figure~\ref{fig:rho_distribution}.
We see for primary vehicle jobs on average 24\% of a job time is location
dependent, while it is 36.5\% for secondary vehicles. Considering
location-dependent service rates would be an interesting further research
question, potentially improving the model's performance by incorporating travel
times into the MESLMHPHF. However, given that in the case of Jakarta the
preferences $\beta_{p a_1 a_2}$ and $R_{p a_1 a_2}$ are themselves defined by
travel times (Equations~\ref{eqn:beta} and~\ref{eqn:R}), this might not have a strong impact on the results of \st{may not have much
affect for} this study, as the model already gives preference to
the locations with the shortest service times.

@@

We propose finding the true vehicle utilisations by first solving
Equation~\ref{eqn:utilisations_primary} for the $\lambda_a$; determining the
primary vehicle utilisations with Equation~\ref{eqn:utilisation_ratio_primary};
then using these to solve Equation~\ref{eqn:utilisations_secondary} for the
$\tilde{\lambda}_a$; and determining the secondary vehicle utilisations with
Equation~\ref{eqn:utilisation_ratio_secondary}. These can be solved
numerically. In our implementation this is solved using the MINPACK hybrd and
hybrj algorithms by using Scipy's \texttt{fsolve} function
\citep{2020SciPy-NMeth}. These algorithms are efficient methods for finding zeros of systems of non-linear equations. 
They are based on the Powell hybrid method \citep{powell70}. It should be noted that in this section we assume that
all $\lambda_{pk}$ are static and not time-dependant, which is an assumption
that will be used throughout the optimisation methodology, however in the
simulation we can account for time-dependent demand.

@@

\begin{algorithm}
\algnewcommand\algorithmicto{\textbf{to }}
\caption{Evolutionary Algorithm used to find better allocations.}
\label{alg:heuristic}
\begin{algorithmic}[1]
\State Create a population of $N$ random allocations ($Z_a$, $\tilde{Z}_a$)\;
\For{$i \leftarrow 0$ \algorithmicto $H$}
  \State Rank the population by decreasing expected survival according to Equation~\ref{eqn:meslmhphf}\;
  \State Keep the top $\kappa$ solutions\;
  \State $m \leftarrow \lceil m_0 - \min\left(m_0, i \times c\right)\rceil$\;
  \For{$j \leftarrow 0$ \algorithmicto $N - \kappa$}
    \State Randomly choose and copy a solution from the population\;
    \State Mutate that copy of the solution $m$ times\;
    \State Place mutated copy in a separate population\;
  \EndFor
  \State Combine populations;\
\EndFor
\State Rank the population according to Equation~\ref{eqn:meslmhphf};\
\State Output the top ranking solution.
\end{algorithmic}
\end{algorithm}

@@

This is the method used to calculate the expected values of $B_{pa}$, $C_{py}$,
$D_{ya}$ and $F_{pa}$.
Furthermore it is assumed that travel times for each segment of the route
(from vehicle location to patient, from patient to hospital, and from hospital
back the the vehicle's location) follow a Triangular distribution around the
calculated expected travel time, between 75\% and 125\% of the value.
A Triangular distribution is a useful distribution to use when
not much is known about the travel times except some basic summary statistics
\citep{robinson2014simulation}, in this case a likely range of values.

@@

This is possible as there are no synchronicity issues between the two
components. Primary vehicles operate independently of secondary vehicles, that
is, the way primary vehicles respond to a patient is unaffected by the presence
or lack of a secondary vehicle.
Therefore, primary vehicle logic is not compromised by simulating primary
vehicles in isolation.
Secondary vehicles are impacted by the behaviour of primary vehicles, they
must remain with the patient until the primary vehicle arrives, and so must be
simulated after the simulation of primary vehicles, taking as inputs the exact
list of events that occurred. That is, the logic of the secondary vehicles is
determined by observing the actions of primary vehicles and reacting to them.
Simulation results are combined for each individual patient, to determine
their response time caused by either primary or secondary vehicles.

@@

Exact synchronicity considerations are shown in Figure~\ref{fig:sequential_logic}:
\begin{itemize}
  \item whenever the secondary vehicle reaches the patient before the primary
  has left, vehicle, they remain with the patient until the primary vehicle
  leaves;
  \item if secondary vehicle arrives after the primary vehicle has left, they
  will immediately return to their stations as they are not needed at the scene;
  \item if the expected time for the secondary vehicle to reach the patient
  exceeds the expected time for the primary vehicle to reach the patient then
  the secondary vehicle is not deployed, and so that transit job would be
  abandoned in the secondary simulation (although still seen by a primary
  vehicle in the initial simulation). We assume that secondary vehicles must
  still reach the pickup locations, due to potential difficulties in
  communication en-route. Consider a situation when a
  secondary vehicle arrives and the patient does not require transportation to
  a hospital, with better en-route communication this would mean the primary
  vehicle turns around, however without en-route communication which is
  assumed here, the primary vehicle must still reach the patient pick-up
  location.
\end{itemize}

@@

Verification of a model is ensuring that the conceptual model is implemented
correctly and without error, while validation of a model ensures that the
conceptual model accurately represents the real system to be modelled
\citep{robinson2014simulation}. Verifying the simulation code, alongside
verifying the optimisation and MESLMHPHF objective function code, is done
through automated code testing \citep{percival2014test, palmer2023implementing}.
This includes unit testing to ensure that each sub-component works are requires,
and is bug and error free; end to end testing, ensuring that the components
interact together as intended; and extreme value testing ensuring that changes
in input parameters result in sensible and expected changes in output results
and behaviours. The testing suites are included in the open code repositories
provided.

@@

\begin{itemize}
  \item \textbf{A1 - High priority emergency patients}: critical patients who
  require immediate life-saving assistance with a target ambulance response
  time of 8 minutes. In our data set there were 168 identified calls that met
        this criterion. \textcolor{red}{As in~\cite{Knight2012918},
        equation~(\ref{eqn:survival_curve}) is here used to measure survival of these patients.}
  \item \textbf{A2 - Other emergency patients}: urgent patients that require
  assistance with a target response time of 15 minutes. In our data set there
  were 23,784 calls of this type.
  \item \textbf{B - Non-emergency patients}: patients that still benefit from
  an ambulance response but non-critical with a target response time of 60
  minutes. There were 31,725 calls of this type in the data set.
\end{itemize}

@@

All models are parameterised using demand data that covers all 261
neighbourhoods from 1 January to 31 December 2019, before the COVID-19
pandemic. Data from 2020-2021 was naturally heavily skewed by the pandemic,
with significantly lower demand, and so was not considered as representative
of a typical period for forecasting future needs, hence the decision to use
the available 2019 demand. APPENDIX~\ref{apx:parameters}
provides further details on the parameterisation of the model, including how
travel time estimates were obtained.
Verification of the models were discussed in
Section~\ref{sec:combined_model}, we validated based on expert opinion from
our collaborators, Ambulans 118 and the 119 Emergency Ambulance Service in
Jakarta, where several workshops have been carried out in Indonesia where the
model was shared and discussed. This expertise was vital in giving a basic
validation of the model, with collaborators approving model results
parametrised from their provided data and expert opinion.

@@

Running the simulation and MESLMHPHF objective function of expected survival for
both the current and the two proposed grid allocations gives the results shown
in Table~\ref{tbl:current_grid_results}. The simulation was run over half a
year, with a month warm up time, over eight replications. 
\textcolor{red}{The variability over the eight replications ...}
We observe that the
two proposed grid allocations perform much worse than the current allocation as
measured by mean response time, survival, and percentage of calls abandoned.
This is expected as vehicle numbers are lower under the new proposals, as well
as the allocations focussing on coverage rather than response times or survival.
